---
title: "Grey Model — Outbreak Comparative Analysis"
author: "Andy Chu"
date: "`r Sys.Date()`"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "d:/RA/Grey model")
libraries <- c(
  "tidyverse", "readxl", "kableExtra", "GreyModel", "Greymodels"
)
for (x in libraries) {
  suppressMessages(
    library(x, character.only = TRUE,
            warn.conflicts = FALSE, quietly = TRUE)
  )
}
```

## Functions {-}

### Calculate Errors, mae() and mape() {-}
```{r fn-errors}
mae <- function(actual, predicted) {
  mean(abs(actual - predicted), na.rm = TRUE)
}

mape <- function(actual, predicted) {
  valid <- actual != 0
  mean(
    abs((actual[valid] - predicted[valid]) / actual[valid]),
    na.rm = TRUE
  ) * 100
}
```

### Grey Model with Irregular Time Series, gmits() {-}
```{r fn-gmits}
# n_ahead: out-of-sample steps to forecast using mean(delta_t) as step size.
# Returns $x0_future when n_ahead > 0.
gmits <- function(count, date, n_ahead = 0) {
  valid <- !is.na(count) & !is.na(date)
  x0    <- count[valid]
  date  <- date[valid]
  if (length(x0) < 2) {
    warning("Not enough valid data points for modeling.")
    return(NULL)
  }
  if (x0[1] == 0) x0[1] <- 1e-3
  n <- length(x0)

  delta_t    <- as.numeric(
    difftime(date[2:n], date[1:(n - 1)], units = "days")
  )
  time_index <- cumsum(c(0, delta_t))

  # Time-weighted AGO: first point as anchor, rest scaled by interval
  x1 <- cumsum(c(x0[1], x0[2:n] * delta_t))
  z  <- 0.5 * (x1[2:n] + x1[1:(n - 1)])

  B <- cbind(-z * delta_t, delta_t)
  Y <- x0[2:n]
  W <- diag(delta_t)
  p <- solve(t(B) %*% W %*% B) %*% t(B) %*% W %*% Y
  a_hat <- p[1]
  b_hat <- p[2]

  # Analytical solution: x1 trajectory, then IAGO back to period counts
  x1_hat <- c(
    x1[1],
    (x1[1] - b_hat / a_hat) *
      exp(-a_hat * (time_index[-1] - time_index[1])) +
      b_hat / a_hat
  )
  x0_hat <- c(x0[1], diff(x1_hat) / delta_t)

  result <- list(
    a = a_hat, b = b_hat,
    x1_hat = x1_hat, x0_hat = x0_hat
  )

  # Out-of-sample forecast: step forward using mean reporting interval
  if (n_ahead > 0) {
    dt_mean  <- mean(delta_t)
    t_future <- time_index[n] + cumsum(rep(dt_mean, n_ahead))
    x1_fut   <- (x1[1] - b_hat / a_hat) *
      exp(-a_hat * (t_future - time_index[1])) + b_hat / a_hat
    result$x0_future <- c(
      (x1_fut[1] - x1_hat[n]) / dt_mean,
      diff(x1_fut) / dt_mean
    )
  }
  result
}
```

### Non-homogeneous Discrete Grey Model, ndgm() {-}
```{r fn-ndgm}
ndgm <- function(count, date, ahead = 4) {
  x0 <- count
  if (x0[1] == 0) x0[1] <- 1e-3
  x1 <- cumsum(x0)
  n  <- length(x0)

  delta_t    <- as.numeric(
    difftime(date[2:n], date[1:(n - 1)], units = "days")
  )
  time_index <- cumsum(c(0, delta_t))

  B    <- cbind(
    matrix(x1[1:(n - 1)], ncol = 1),
    matrix(time_index[1:(n - 1)], ncol = 1),
    matrix(1, nrow = n - 1, ncol = 1)
  )
  beta <- solve(t(B) %*% B) %*% t(B) %*% matrix(x1[2:n], ncol = 1)
  b1   <- beta[1, 1]
  b2   <- beta[2, 1]
  b3   <- beta[3, 1]

  k_all  <- c(
    time_index,
    max(time_index) + cumsum(rep(mean(delta_t), ahead))
  )
  x1cap  <- numeric(length(k_all))
  for (i in seq_along(k_all)) {
    k        <- k_all[i]
    sm       <- sum((1:i) * b1^(k - k_all[1:i]))
    x1cap[i] <- b1^k * x1[1] +
      b2 * sm + ((1 - b1^k) / (1 - b1)) * b3
  }
  c(x0[1], diff(x1cap))
}
```

### Right Truncation Adjustment, truncation_adjust() {-}
```{r fn-truncation}
# Scales the most recent n_recent counts by scale_factor before passing to gmits().
truncation_adjust <- function(count, n_recent = 1, scale_factor = 1.5) {
  n   <- length(count)
  adj <- count
  idx <- tail(seq_len(n), n_recent)
  adj[idx] <- count[idx] * scale_factor
  adj
}
```

### Bootstrap Prediction Interval, gmits_boot() {-}
```{r fn-bootstrap}
# n_ahead > 0: point/lower/upper each extend to n + n_ahead;
# last n_ahead entries are the out-of-sample prediction interval.
gmits_boot <- function(count, date, n_boot = 1000, ci = 0.95, n_ahead = 0) {
  base <- gmits(count, date, n_ahead = n_ahead)
  if (is.null(base)) return(NULL)

  valid      <- !is.na(count) & !is.na(date)
  x0         <- count[valid]
  dv         <- date[valid]
  n          <- length(x0)
  resid_pool <- (x0 - base$x0_hat)[2:n]
  n_out      <- n + n_ahead

  boot_mat <- replicate(n_boot, {
    e_boot  <- sample(resid_pool, n - 1, replace = TRUE)
    x0_boot <- c(x0[1], pmax(base$x0_hat[2:n] + e_boot, 0))
    res     <- tryCatch(
      gmits(x0_boot, dv, n_ahead = n_ahead),
      error = function(e) NULL
    )
    if (is.null(res)) return(rep(NA_real_, n_out))
    c(res$x0_hat, if (n_ahead > 0) res$x0_future else NULL)
  })

  alpha <- (1 - ci) / 2
  list(
    point = c(base$x0_hat,
              if (n_ahead > 0) base$x0_future else NULL),
    lower = apply(boot_mat, 1, quantile, probs = alpha,     na.rm = TRUE),
    upper = apply(boot_mat, 1, quantile, probs = 1 - alpha, na.rm = TRUE),
    a = base$a,
    b = base$b
  )
}
```

---

# Data

## Bongkrek Acid Poisoning (Point-Source, 2024)

6 exposure dates, 17 linelist snapshots (Mar 24 – Apr 7).
Point-source food poisoning; cases are not transmissible.

```{r data-bongkrek}
df_bongkrek <- data.frame(
  Date           = as.Date(c(
    "2024/3/18", "2024/3/19", "2024/3/21",
    "2024/3/22", "2024/3/23", "2024/3/24"
  )),
  Tracking_4_07b = c(1, 5, 6, 8, 6, 8),
  Tracking_4_06b = c(1, 5, 6, 8, 6, 8),
  Tracking_4_05b = c(1, 5, 6, 8, 6, 8),
  Tracking_4_04b = c(0, 5, 6, 8, 6, 8),
  Tracking_4_03b = c(0, 5, 6, 7, 6, 8),
  Tracking_4_02b = c(0, 5, 6, 7, 6, 6),
  Tracking_4_01a = c(0, 5, 6, 7, 5, 6),
  Tracking_3_31b = c(0, 5, 6, 7, 5, 6),
  Tracking_3_31a = c(0, 5, 5, 7, 5, 6),
  Tracking_3_30b = c(0, 4, 4, 7, 5, 5),
  Tracking_3_30a = c(0, 4, 4, 7, 5, 4),
  Tracking_3_29b = c(0, 2, 4, 6, 5, 4),
  Tracking_3_29a = c(0, 2, 4, 4, 4, 4),
  Tracking_3_28a = c(0, 2, 4, 4, 4, 0),
  Tracking_3_27b = c(0, 2, 3, 3, 3, 0),
  Tracking_3_26b = c(0, 2, 2, 2, 3, 0),
  Tracking_3_24b = c(0, 0, 0, 2, 0, 0)
)

snapshot_cols <- grep("^Tracking", names(df_bongkrek), value = TRUE)
# Ordered from earliest to latest tracking date
snapshot_cols <- rev(snapshot_cols)
```

```{r plot-bongkrek-epicurve}
df_bongkrek |>
  select(Date, Tracking_4_07b) |>
  rename(Case = Tracking_4_07b) |>
  ggplot(aes(Date, Case)) +
  geom_col(fill = "steelblue") +
  scale_x_date(date_labels = "%m-%d", date_breaks = "1 day") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Bongkrek Acid — Final Epicurve (Apr 7 snapshot)",
       x = "Date", y = "Cases")
```

## 2011 Listeria Outbreak (Ongoing Transmission)

Multiple weekly snapshots; ongoing foodborne transmission (person-to-person spread possible).
Three representative snapshots are selected by cumulative case percentage of the final total (targets: ~20%, 50%, 80%).

```{r data-listeria}
wd <- "d:/RA/Listeria/data mdy"
files <- list.files(wd, pattern = "\\.csv$", full.names = FALSE)

listeria <- map_df(
  files,
  ~ read_csv(
      file.path(wd, .x),
      show_col_types = FALSE
    ) |>
    mutate(Snapshot = str_remove(.x, "\\.csv$"))
) |>
  pivot_wider(
    id_cols     = Date,
    names_from  = Snapshot,
    values_from = Count
  ) |>
  mutate(Date = as.Date(Date))

# Select snapshots by cumulative case % of final total
snapshot_totals <- listeria |>
  select(-Date) |>
  summarise(across(everything(), \(x) sum(x, na.rm = TRUE))) |>
  pivot_longer(
    everything(),
    names_to  = "Snapshot",
    values_to = "Total"
  ) |>
  mutate(Pct = Total / max(Total) * 100)

pick_snap <- function(target) {
  snapshot_totals$Snapshot[
    which.min(abs(snapshot_totals$Pct - target))
  ]
}
snap_beginning <- pick_snap(20)
snap_middle    <- pick_snap(50)
snap_end       <- pick_snap(80)

make_phase <- function(snap, label) {
  listeria |>
    select(Date, all_of(snap)) |>
    rename_with(\(x) "Case", all_of(snap)) |>
    drop_na() |>
    mutate(Phase = label)
}
lst_beginning <- make_phase(snap_beginning, "Beginning")
lst_middle    <- make_phase(snap_middle,    "Middle")
lst_end       <- make_phase(snap_end,       "End")

lst_combined <- bind_rows(lst_beginning, lst_middle, lst_end) |>
  mutate(
    Phase = factor(Phase, levels = c("Beginning", "Middle", "End")),
    Date  = as.Date(Date)
  )
```

```{r listeria-phase-table}
snapshot_totals |>
  filter(Snapshot %in% c(snap_beginning, snap_middle, snap_end)) |>
  mutate(
    Phase = case_when(
      Snapshot == snap_beginning ~ "Beginning",
      Snapshot == snap_middle    ~ "Middle",
      Snapshot == snap_end       ~ "End"
    ),
    Phase = factor(Phase, levels = c("Beginning", "Middle", "End")),
    Pct   = round(Pct, 1)
  ) |>
  arrange(Phase) |>
  select(Phase, Snapshot, Total, Pct) |>
  kable(
    col.names = c("Phase", "Snapshot", "Total Cases", "% of Final"),
    caption   = "Selected Listeria Snapshots by Cumulative Case %"
  )
```

```{r plot-listeria-epicurve}
phase_colors <- c(
  Beginning = "#1b9e77",
  Middle    = "#d95f02",
  End       = "#7570b3"
)

ggplot(lst_combined, aes(Date, Case, fill = Phase)) +
  geom_col(alpha = 0.6, position = "identity") +
  scale_fill_manual(values = phase_colors) +
  scale_x_date(date_labels = "%m-%d", date_breaks = "7 days") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Listeria 2011 — Epicurve by Snapshot Phase",
       x = "Date", y = "Cases")
```

---

# Synthetic Validation — GM(1,1) Baseline

Validates GM(1,1) behaviour on synthetic exponential-growth data with known parameters and added noise.

```{r synthetic-data}
set.seed(5656)
a_true  <- 0.3
n_obs   <- 5
t_seq   <- seq(1, n_obs)
s_true  <- exp(a_true * (t_seq - 1))
x_obs   <- s_true + rnorm(n_obs, 0, sd = sqrt(0.1))
snr     <- var(s_true) / var(x_obs - s_true)

df_synth <- data.frame(
  true_value  = s_true,
  observation = x_obs
) |>
  add_row(
    true_value  = rep(NA, 4),
    observation = rep(NA, 4)
  ) |>
  mutate(
    prediction = gm11(x_obs),
    time_step  = seq_len(n_obs + 4)
  )
```

```{r plot-synthetic}
df_synth |>
  pivot_longer(
    c(observation, prediction, true_value),
    names_to  = "type",
    values_to = "value"
  ) |>
  ggplot(aes(time_step, value, color = type)) +
  geom_line() +
  geom_vline(xintercept = n_obs + 0.5, linetype = "dashed",
             color = "grey50") +
  annotate("text", x = n_obs + 0.6, y = max(df_synth$prediction,
           na.rm = TRUE) * 0.9,
           label = "forecast", hjust = 0, size = 3) +
  theme_minimal() +
  labs(
    title = paste0("GM(1,1) on Synthetic Exponential Data (SNR = ",
                   round(snr, 1), ")"),
    x = "Time step", y = "Count", color = ""
  )
```

```{r error-synthetic}
data.frame(
  MAE  = mae(df_synth$true_value, df_synth$prediction),
  MAPE = mape(df_synth$true_value, df_synth$prediction)
) |>
  kable(digits = 3, caption = "GM(1,1) in-sample error (synthetic)")
```

---

# NDGM — Real-time Forecast Convergence

Each tracking-date snapshot is modelled independently with a 4-step-ahead forecast,
illustrating how predictions converge as data accumulate.

## Bongkrek: 17 Snapshots

```{r ndgm-bongkrek}
ndgm_list <- lapply(snapshot_cols, function(col) {
  pred <- ndgm(df_bongkrek[[col]], df_bongkrek$Date, ahead = 4)
  data.frame(
    Predicted = pred,
    Snapshot  = col,
    Date      = seq.Date(
      min(df_bongkrek$Date),
      by = "day",
      length.out = length(pred)
    )
  )
})
df_ndgm_bongkrek <- bind_rows(ndgm_list)
```

```{r plot-ndgm-bongkrek}
ggplot() +
  geom_col(
    data = df_bongkrek,
    aes(Date, Tracking_4_07b),
    fill = "grey80"
  ) +
  geom_line(
    data = df_ndgm_bongkrek,
    aes(Date, Predicted, color = Snapshot)
  ) +
  scale_x_date(date_labels = "%m-%d", date_breaks = "1 day") +
  guides(color = guide_legend(ncol = 2)) +
  theme_minimal() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    legend.text  = element_text(size = 7),
    legend.key.size = unit(0.4, "cm")
  ) +
  labs(
    title = "NDGM — Real-time Forecast Convergence",
    x = "Date", y = "Count", color = "Snapshot"
  )
```

---

# GMITS — Comparative Fit by Phase

GMITS is fitted independently to each snapshot; error metrics are compared across epidemic stages.

## Bongkrek: Each Snapshot

```{r gmits-bongkrek}
# Bongkrek snapshots encode unreported cases as 0, not NA.
# For sequential forecasting, 0 means "not yet attributed in this snapshot"
# and must be treated as missing — otherwise gmits() fits through spurious zeros.
# Snapshots with fewer than 4 valid (non-zero) observations are skipped.
gmits_bongkrek <- lapply(snapshot_cols, function(col) {
  obs_raw <- df_bongkrek[[col]]
  obs     <- na_if(obs_raw, 0L)          # 0 → NA: not yet attributed
  if (sum(!is.na(obs)) < 4) return(NULL) # need ≥4 points for GM(1,1)
  res <- gmits(obs, df_bongkrek$Date)
  if (is.null(res)) return(NULL)
  valid <- !is.na(obs)
  data.frame(
    Snapshot = col,
    n_valid  = sum(valid),
    MAE      = mae(obs[valid], res$x0_hat),
    MAPE     = mape(obs[valid], res$x0_hat),
    a        = res$a,
    b        = res$b
  )
}) |>
  bind_rows()
```

```{r plot-gmits-bongkrek}
# Final snapshot has no zeros; na_if() is a no-op here but kept for consistency.
final_snap <- "Tracking_4_07b"
obs_final  <- na_if(df_bongkrek[[final_snap]], 0L)
res_final  <- gmits(obs_final, df_bongkrek$Date)

data.frame(
  Date      = df_bongkrek$Date[!is.na(obs_final)],
  Observed  = obs_final[!is.na(obs_final)],
  Predicted = res_final$x0_hat
) |>
  pivot_longer(c(Observed, Predicted),
               names_to = "type", values_to = "value") |>
  ggplot(aes(Date, value, color = type)) +
  geom_line() +
  geom_point() +
  scale_x_date(date_labels = "%m-%d", date_breaks = "1 day") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "GMITS Fit — Bongkrek (final snapshot)",
    x = "Date", y = "Cases", color = ""
  )
```

## Listeria: Three Phases

```{r gmits-listeria}
lst_phases   <- list(
  Beginning = lst_beginning,
  Middle    = lst_middle,
  End       = lst_end
)
phase_levels <- c("Beginning", "Middle", "End")

# Fit once; extract error metrics and fitted values from the same run.
listeria_fits <- lapply(names(lst_phases), function(phase) {
  df  <- lst_phases[[phase]]
  res <- gmits(df$Case, df$Date)
  if (is.null(res)) return(NULL)
  list(
    errors = data.frame(
      Phase    = phase,
      n_points = nrow(df),
      MAE      = mae(df$Case, res$x0_hat),
      MAPE     = mape(df$Case, res$x0_hat),
      a        = res$a,
      b        = res$b
    ),
    fit = data.frame(
      Date      = df$Date,
      Observed  = df$Case,
      Predicted = res$x0_hat,
      Phase     = phase
    )
  )
})

gmits_listeria <- bind_rows(lapply(listeria_fits, `[[`, "errors")) |>
  mutate(Phase = factor(Phase, levels = phase_levels))

df_lst_fit <- bind_rows(lapply(listeria_fits, `[[`, "fit")) |>
  mutate(Phase = factor(Phase, levels = phase_levels))
```

```{r plot-gmits-listeria}
ggplot(df_lst_fit, aes(Date)) +
  geom_col(aes(y = Observed, fill = Phase), alpha = 0.4) +
  geom_line(aes(y = Predicted, color = Phase), linewidth = 0.8) +
  scale_fill_manual(values = phase_colors) +
  scale_color_manual(values = phase_colors) +
  scale_x_date(date_labels = "%m-%d", date_breaks = "7 days") +
  facet_wrap(~Phase, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "GMITS Fit — Listeria by Phase",
    x = "Date", y = "Cases"
  )
```

## Listeria: Bootstrap 95% Prediction Interval (n_ahead = 4)

```{r gmits-listeria-boot}
# n_ahead = 4: produce prediction interval for 4 steps beyond each snapshot.
# The first n rows are in-sample; the last 4 are the out-of-sample forecast PI.
n_fwd <- 4
boot_list <- lapply(names(lst_phases), function(phase) {
  df  <- lst_phases[[phase]]
  res <- gmits_boot(df$Case, df$Date, n_boot = 1000, n_ahead = n_fwd)
  if (is.null(res)) return(NULL)
  n_obs   <- nrow(df)
  dt_mean <- mean(as.numeric(diff(df$Date)))
  data.frame(
    Date     = c(df$Date,
                 df$Date[n_obs] + round(dt_mean * seq_len(n_fwd))),
    Point    = res$point,
    Lower    = res$lower,
    Upper    = res$upper,
    Forecast = c(rep(FALSE, n_obs), rep(TRUE, n_fwd)),
    Phase    = phase
  )
})
df_boot <- bind_rows(boot_list) |>
  mutate(Phase = factor(Phase, levels = c("Beginning", "Middle", "End")))
```

```{r plot-gmits-listeria-boot}
ggplot(df_boot, aes(Date)) +
  geom_col(
    data  = df_lst_fit |> select(Date, Observed, Phase),
    aes(y = Observed, fill = Phase),
    alpha = 0.35
  ) +
  geom_ribbon(
    aes(ymin = Lower, ymax = Upper, fill = Phase, linetype = Forecast),
    alpha = 0.3
  ) +
  geom_line(
    aes(y = Point, color = Phase, linetype = Forecast),
    linewidth = 0.8
  ) +
  scale_fill_manual(values  = phase_colors) +
  scale_color_manual(values = phase_colors) +
  scale_linetype_manual(
    values = c("FALSE" = "solid", "TRUE" = "dashed"),
    labels = c("FALSE" = "in-sample", "TRUE" = "forecast")
  ) +
  scale_x_date(date_labels = "%m-%d", date_breaks = "7 days") +
  facet_wrap(~Phase, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "GMITS Bootstrap 95% PI — Listeria by Phase (dashed = forecast)",
    x = "Date", y = "Cases", linetype = ""
  )
```

## Bongkrek: Bootstrap 95% Prediction Interval

```{r gmits-bongkrek-boot}
# Final (most complete) Bongkrek snapshot; forecast 3 daily steps ahead.
obs_boot <- na_if(df_bongkrek[["Tracking_4_07b"]], 0L)
dates_b  <- df_bongkrek$Date[!is.na(obs_boot)]
n_obs_b  <- length(dates_b)
dt_b     <- mean(as.numeric(diff(dates_b)))

res_bong_boot <- gmits_boot(
  obs_boot, df_bongkrek$Date,
  n_boot = 1000, n_ahead = 3
)

df_bong_boot <- data.frame(
  Date     = c(dates_b,
               tail(dates_b, 1) + round(dt_b * seq_len(3))),
  Point    = res_bong_boot$point,
  Lower    = res_bong_boot$lower,
  Upper    = res_bong_boot$upper,
  Forecast = c(rep(FALSE, n_obs_b), rep(TRUE, 3))
)
```

```{r plot-gmits-bongkrek-boot}
ggplot(df_bong_boot, aes(Date)) +
  geom_col(
    data = df_bongkrek,
    aes(y = Tracking_4_07b),
    fill = "grey75", alpha = 0.6
  ) +
  geom_ribbon(
    aes(ymin = Lower, ymax = Upper, linetype = Forecast),
    fill = "steelblue", alpha = 0.3
  ) +
  geom_line(
    aes(y = Point, linetype = Forecast),
    color = "steelblue", linewidth = 0.9
  ) +
  scale_linetype_manual(
    values = c("FALSE" = "solid", "TRUE" = "dashed"),
    labels = c("FALSE" = "in-sample", "TRUE" = "forecast")
  ) +
  scale_x_date(date_labels = "%m-%d", date_breaks = "1 day") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "GMITS Bootstrap 95% PI — Bongkrek (dashed = forecast)",
    x = "Date", y = "Cases", linetype = ""
  )
```

---

# Right Truncation Sensitivity

Compares GMITS predictions under different right truncation assumptions, where recent counts
are systematically underreported. `truncation_adjust()` multiplies the latest `n_recent`
observations by `scale_factor` as a sensitivity pre-processing step.

```{r rt-sensitivity}
# Compare GMITS fit under three truncation assumptions for Listeria Beginning phase.
df_rt <- lst_beginning
scales <- c(none = 1.0, mild = 1.5, moderate = 2.0)

rt_results <- lapply(names(scales), function(nm) {
  adj   <- truncation_adjust(df_rt$Case, n_recent = 1,
                             scale_factor = scales[[nm]])
  res   <- gmits(adj, df_rt$Date, n_ahead = 4)
  if (is.null(res)) return(NULL)
  n_obs <- nrow(df_rt)
  dt_rt <- mean(as.numeric(diff(df_rt$Date)))
  data.frame(
    Date        = c(df_rt$Date,
                    df_rt$Date[n_obs] + round(dt_rt) * (1:4)),
    Predicted   = c(res$x0_hat, res$x0_future),
    Assumption  = nm,
    Forecast    = c(rep(FALSE, n_obs), rep(TRUE, 4))
  )
})
df_rt_all <- bind_rows(rt_results) |>
  mutate(Assumption = factor(Assumption, levels = names(scales)))
```

```{r plot-rt-sensitivity}
ggplot(df_rt_all, aes(Date, Predicted,
                      color = Assumption, linetype = Forecast)) +
  geom_line(linewidth = 0.8) +
  geom_col(
    data  = df_rt,
    aes(Date, Case),
    inherit.aes = FALSE,
    fill = "grey75", alpha = 0.5
  ) +
  scale_linetype_manual(
    values = c("FALSE" = "solid", "TRUE" = "dashed"),
    labels = c("FALSE" = "in-sample", "TRUE" = "forecast")
  ) +
  scale_x_date(date_labels = "%m-%d", date_breaks = "7 days") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Right Truncation Sensitivity — Listeria Beginning Phase",
    subtitle = "none = no adjustment; mild = ×1.5; moderate = ×2.0 on latest point",
    x = "Date", y = "Cases", color = "Assumption", linetype = ""
  )
```

---

# Error Summary

```{r error-summary}
bind_rows(
  gmits_bongkrek |>
    filter(Snapshot %in% c(
      "Tracking_3_26b", "Tracking_3_29b",
      "Tracking_4_03b", "Tracking_4_07b"
    )) |>
    mutate(Dataset = "Bongkrek", Phase = Snapshot) |>
    select(Dataset, Phase, MAE, MAPE, a, b),
  gmits_listeria |>
    mutate(
      Dataset = "Listeria",
      Phase   = as.character(Phase)
    ) |>
    select(Dataset, Phase, MAE, MAPE, a, b)
) |>
  kable(
    digits  = 3,
    caption = "GMITS In-Sample Error by Dataset and Phase"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover")) |>
  pack_rows("Bongkrek (snapshot evolution)", 1, 4) |>
  pack_rows("Listeria 2011 (outbreak phase)", 5, 7)
```
