---
title: "HFRS Incubation Period Estimation"
subtitle: "Kulagin et al. (1962) — Laboratory Outbreak, n = 18"
date: "`r Sys.Date()`"
author: "Andy Chu"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

------------------------------------------------------------------------

## 1. Overview

Incubation periods were recorded from 18 laboratory workers (Group A:
single-visit external visitors who attended the department on 24 Oct 1961
only) who visited the infected department **exactly once**, providing a
single point-identified exposure date.

Because both the exposure date and the symptom-onset date were only
recorded to the calendar day, the true continuous incubation period is
never observed exactly — it is **doubly interval-censored**. Each
recorded integer $k$ represents a true incubation $T \in (k - 1,\, k + 1)$.
This motivates an interval-censored likelihood rather than treating the
integer counts as exact point observations (used in Section 3).

Source: Kulagin, Fedorova & Ketiladze (1962), *Journal of Microbiology,
Epidemiology and Immunology*, 33(10), 121–126.
<https://www.cabidigitallibrary.org/doi/full/10.5555/19632701583>

```{r data}
inc <- c(rep(17, 3), 19, rep(21, 2), rep(22, 8), 23, rep(25, 2), 27)
cat("n =", length(inc), "\n")
cat("Range:", min(inc), "-", max(inc), "days\n")
cat("Mean:", round(mean(inc), 1), "  Median:", median(inc), "\n")
```

```{r data-plot, fig.width=5, fig.height=4}
ggplot(data.frame(x = inc), aes(x = "", y = x)) +
  geom_boxplot(
    fill = "#009E73", alpha = 0.4, width = 0.4, outlier.shape = NA
  ) +
  geom_jitter(width = 0.08, alpha = 0.6, color = "#009E73", size = 2) +
  labs(
    title    = "Observed Incubation Periods (n = 18, Group A)",
    subtitle = paste0(
      "Single-visit visitors on Oct 24; ",
      "integer-day difference between exposure and onset"
    ),
    x = NULL, y = "days"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## 2. Maximum Likelihood Estimation

### 2.1 Fitting

Three parametric families are fitted via MLE with L-BFGS-B optimization.
These use point observations for comparison; the interval-censored
treatment is applied in Section 3.

```{r mle, results='hide'}
ll_lnorm <- function(params, data) {
  mu    <- params[1]
  sigma <- params[2]
  if (sigma <= 0) return(Inf)
  -sum(dlnorm(data, meanlog = mu, sdlog = sigma, log = TRUE))
}
fit_lnorm <- optim(
  c(mean(log(inc)), sd(log(inc))), ll_lnorm, data = inc,
  method = "L-BFGS-B", lower = c(-Inf, 0.001), hessian = TRUE
)
se_lnorm <- sqrt(diag(solve(fit_lnorm$hessian)))
ci_lnorm <- cbind(
  fit_lnorm$par - 1.96 * se_lnorm,
  fit_lnorm$par + 1.96 * se_lnorm
)
aic_lnorm <- 2 * 2 + 2 * fit_lnorm$value

ll_weib <- function(params, data) {
  shape <- params[1]
  scale <- params[2]
  if (shape <= 0 || scale <= 0) return(Inf)
  -sum(dweibull(data, shape = shape, scale = scale, log = TRUE))
}
fit_weib <- optim(
  c(1, mean(inc)), ll_weib, data = inc,
  method = "L-BFGS-B", lower = c(0.001, 0.001), hessian = TRUE
)
se_weib <- sqrt(diag(solve(fit_weib$hessian)))
ci_weib <- cbind(
  fit_weib$par - 1.96 * se_weib,
  fit_weib$par + 1.96 * se_weib
)
aic_weib <- 2 * 2 + 2 * fit_weib$value

ll_gamma <- function(params, data) {
  shape <- params[1]
  rate  <- params[2]
  if (shape <= 0 || rate <= 0) return(Inf)
  -sum(dgamma(data, shape = shape, rate = rate, log = TRUE))
}
fit_gamma <- optim(
  c(1, 1 / mean(inc)), ll_gamma, data = inc,
  method = "L-BFGS-B", lower = c(0.001, 0.001), hessian = TRUE
)
se_gamma <- sqrt(diag(solve(fit_gamma$hessian)))
ci_gamma <- cbind(
  fit_gamma$par - 1.96 * se_gamma,
  fit_gamma$par + 1.96 * se_gamma
)
aic_gamma <- 2 * 2 + 2 * fit_gamma$value
```

### 2.2 Parameter Estimates

```{r mle-table}
ci1_lo <- round(
  c(ci_lnorm[1, 1], ci_weib[1, 1], ci_gamma[1, 1]), 3
)
ci1_hi <- round(
  c(ci_lnorm[1, 2], ci_weib[1, 2], ci_gamma[1, 2]), 3
)
ci2_lo <- round(
  c(ci_lnorm[2, 1], ci_weib[2, 1], ci_gamma[2, 1]), 3
)
ci2_hi <- round(
  c(ci_lnorm[2, 2], ci_weib[2, 2], ci_gamma[2, 2]), 3
)
result <- data.frame(
  Distribution = c("Lognormal", "Weibull", "Gamma"),
  Parameters   = c("meanlog, sdlog", "shape, scale", "shape, rate"),
  Param1 = round(
    c(fit_lnorm$par[1], fit_weib$par[1], fit_gamma$par[1]), 4
  ),
  CI1    = paste0("[", ci1_lo, ", ", ci1_hi, "]"),
  Param2 = round(
    c(fit_lnorm$par[2], fit_weib$par[2], fit_gamma$par[2]), 4
  ),
  CI2    = paste0("[", ci2_lo, ", ", ci2_hi, "]"),
  AIC    = round(c(aic_lnorm, aic_weib, aic_gamma), 2)
)
kable(result, align = "llrrrrc",
      col.names = c(
        "Distribution", "Parameters",
        "Param 1", "95% CI", "Param 2", "95% CI", "AIC"
      ))
```

**Interpretation:** AICs fall within ΔAIC < 2 across all three distributions.
Lognormal is retained as the primary model. MLE uses point observations;
the Bayesian model below uses the interval-censored likelihood.

------------------------------------------------------------------------

### 2.3 Distribution Curves (MLE)

```{r mle-plot, fig.width=10, fig.height=8}
freq <- seq(0, 50, length.out = 1000)

df_lnorm <- data.frame(
  x        = freq,
  density  = dlnorm(freq, fit_lnorm$par[1], fit_lnorm$par[2]),
  ci_lower = dlnorm(freq, ci_lnorm[1, 1],   ci_lnorm[2, 1]),
  ci_upper = dlnorm(freq, ci_lnorm[1, 2],   ci_lnorm[2, 2])
)
plnorm <- ggplot(df_lnorm, aes(x = x)) +
  geom_ribbon(
    aes(ymin = ci_lower, ymax = ci_upper),
    fill = "#009E73", alpha = 0.2
  ) +
  geom_line(aes(y = density), color = "#009E73", linewidth = 1) +
  geom_rug(
    data = data.frame(x = inc), aes(x = x),
    inherit.aes = FALSE, sides = "b", alpha = 0.6
  ) +
  labs(
    title = paste0("Lognormal  (AIC = ", round(aic_lnorm, 1), ")"),
    x = "days", y = "density"
  ) +
  theme_minimal()

df_weib <- data.frame(
  x        = freq,
  density  = dweibull(freq, fit_weib$par[1], fit_weib$par[2]),
  ci_lower = dweibull(freq, ci_weib[1, 1],   ci_weib[2, 1]),
  ci_upper = dweibull(freq, ci_weib[1, 2],   ci_weib[2, 2])
)
pweib <- ggplot(df_weib, aes(x = x)) +
  geom_ribbon(
    aes(ymin = ci_lower, ymax = ci_upper),
    fill = "#56B4E9", alpha = 0.2
  ) +
  geom_line(aes(y = density), color = "#56B4E9", linewidth = 1) +
  geom_rug(
    data = data.frame(x = inc), aes(x = x),
    inherit.aes = FALSE, sides = "b", alpha = 0.6
  ) +
  labs(
    title = paste0("Weibull  (AIC = ", round(aic_weib, 1), ")"),
    x = "days", y = "density"
  ) +
  theme_minimal()

df_gamma <- data.frame(
  x        = freq,
  density  = dgamma(freq, fit_gamma$par[1], fit_gamma$par[2]),
  ci_lower = dgamma(freq, ci_gamma[1, 1],   ci_gamma[2, 1]),
  ci_upper = dgamma(freq, ci_gamma[1, 2],   ci_gamma[2, 2])
)
pgamma_p <- ggplot(df_gamma, aes(x = x)) +
  geom_ribbon(
    aes(ymin = ci_lower, ymax = ci_upper),
    fill = "#E69F00", alpha = 0.2
  ) +
  geom_line(aes(y = density), color = "#E69F00", linewidth = 1) +
  geom_rug(
    data = data.frame(x = inc), aes(x = x),
    inherit.aes = FALSE, sides = "b", alpha = 0.6
  ) +
  labs(
    title = paste0("Gamma  (AIC = ", round(aic_gamma, 1), ")"),
    x = "days", y = "density"
  ) +
  theme_minimal()

df_combined <- data.frame(
  x = rep(freq, 3),
  density = c(
    dlnorm(freq,   fit_lnorm$par[1], fit_lnorm$par[2]),
    dweibull(freq, fit_weib$par[1],  fit_weib$par[2]),
    dgamma(freq,   fit_gamma$par[1], fit_gamma$par[2])
  ),
  Distribution = rep(
    c("Lognormal", "Weibull", "Gamma"), each = length(freq)
  )
)
pcom <- ggplot(
  df_combined, aes(x = x, y = density, color = Distribution)
) +
  geom_line(linewidth = 1) +
  geom_rug(
    data = data.frame(x = inc), aes(x = x),
    inherit.aes = FALSE, sides = "b", alpha = 0.6
  ) +
  scale_color_manual(values = c("#009E73", "#E69F00", "#56B4E9")) +
  labs(title = "Combined", x = "days", y = "density") +
  theme_minimal()

grid.arrange(plnorm, pweib, pgamma_p, pcom, ncol = 2)
```

------------------------------------------------------------------------

### 2.4 Q-Q Plot (Lognormal MLE)

```{r qq, fig.width=5, fig.height=5}
n     <- length(inc)
probs <- (seq_len(n) - 0.5) / n
qq_df <- data.frame(
  theoretical = qlnorm(probs, fit_lnorm$par[1], fit_lnorm$par[2]),
  observed    = sort(inc)
)
ggplot(qq_df, aes(x = theoretical, y = observed)) +
  geom_point(color = "#009E73", alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Q-Q Plot — Lognormal (MLE)",
    x = "Theoretical Quantiles", y = "Observed Quantiles"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## 3. Bayesian Estimation via Stan (HMC/NUTS)

### 3.1 Model Specification

#### Distribution Assumption

$$T_{i} \sim \text{LogNormal}(\mu,\, \sigma), \quad i = 1, \ldots, N$$

where $T_{i}$ is the true continuous incubation period of individual $i$
(days).

#### Why exp(μ) ≠ Mean

The lognormal distribution has two distinct location summaries:

$$\text{Median}(T) = \exp(\mu), \qquad
\mathbb{E}[T] = \exp\!\left(\mu + \frac{\sigma^{2}}{2}\right)$$

Reporting $\mu$ directly implies the median, not the mean.
When $\sigma > 0$, the two quantities diverge by a factor of
$\exp(\sigma^{2}/2)$.

#### Reparameterization

Let $m = \mathbb{E}[T]$ (`mean_ip`) and $s = \text{SD}(T)$ (`sd_ip`).
The internal lognormal parameters are recovered by:

$$\mu = \log(m) - \frac{1}{2}\,\log\!\left(1 + \frac{s^{2}}{m^{2}}\right)$$

$$\sigma = \sqrt{\,\log\!\left(1 + \frac{s^{2}}{m^{2}}\right)}$$

Both $m$ and $s$ are in days, allowing priors and credible intervals to
be stated directly in the clinically interpretable scale.

#### Interval-Censored Likelihood

For individual $i$ with exposure window $[E_{L,i},\, E_{R,i}]$ and onset
window $[O_{L,i},\, O_{R,i}]$ (all in calendar days), the continuous
exposure time $E \in (E_{L,i},\; E_{R,i} + 1)$ and onset time
$O \in (O_{L,i},\; O_{R,i} + 1)$, giving per-observation bounds:

$$T^{\text{lo}}_{i} = O_{L,i} - E_{R,i} - 1, \qquad
T^{\text{hi}}_{i} = O_{R,i} - E_{L,i} + 1$$

The log-likelihood contribution is:

$$\ell(\theta \mid i) = \log\!\Bigl[
  F_{\text{LN}}\!\left(T^{\text{hi}}_{i} \mid \mu, \sigma\right)
  - F_{\text{LN}}\!\left(T^{\text{lo}}_{i} \mid \mu, \sigma\right)
\Bigr]$$

Summing over all $N$ observations:

$$\ell(\theta \mid \mathbf{T}) = \sum_{i=1}^{N} \log\!\Bigl[
  F_{\text{LN}}\!\left(T^{\text{hi}}_{i} \mid \mu, \sigma\right)
  - F_{\text{LN}}\!\left(T^{\text{lo}}_{i} \mid \mu, \sigma\right)
\Bigr]$$

**Group A special case** ($E_{L,i} = E_{R,i}$, $O_{L,i} = O_{R,i}$):
$T^{\text{lo}}_{i} = k_{i} - 1$, $T^{\text{hi}}_{i} = k_{i} + 1$ — the
fixed ±1 day window used throughout this analysis. The formulation
generalizes without changes to cases with wider exposure or onset
windows, should such data become available.

```{r stan-model}
stan_code <- "
data {
  int<lower=0> N;
  // Per-case incubation bounds (days, real-valued).
  // For Group A: T_lower = k - 1, T_upper = k + 1
  // (k = integer onset - exposure day).
  // For wider exposure windows: T_lower = O_L - E_R - 1,
  //                             T_upper = O_R - E_L + 1.
  vector<lower=0>[N] T_lower;
  vector<lower=0>[N] T_upper;
}
parameters {
  real<lower=0> mean_ip;
  real<lower=0> sd_ip;
}
transformed parameters {
  real mu    = log(mean_ip)
               - 0.5 * log(1.0 + square(sd_ip / mean_ip));
  real sigma = sqrt(log(1.0 + square(sd_ip / mean_ip)));
}
model {
  // Weakly informative priors -- to be updated after literature review.
  // mean_ip ~ Normal(21, 7): centred near expected HFRS value.
  // sd_ip   ~ HalfNormal(5): <lower=0> enforces positivity.
  mean_ip ~ normal(21, 7);
  sd_ip   ~ normal(0, 5);
  // Interval-censored likelihood.
  for (i in 1:N) {
    target += log_diff_exp(
      lognormal_lcdf(T_upper[i] | mu, sigma),
      lognormal_lcdf(T_lower[i] | mu, sigma)
    );
  }
}
"
```

**Priors:** `mean_ip ~ Normal(21, 7)`, `sd_ip ~ HalfNormal(5)` — weakly
informative, centered near the expected HFRS incubation period pending
literature review.

**Likelihood:** doubly interval-censored; each observation contributes
$P(T \in (T_{\text{lower},i},\, T_{\text{upper},i}))$. For Group A this
is $T_{\text{lower}} = k - 1$, $T_{\text{upper}} = k + 1$. Cases with
wider exposure windows slot in by supplying different bounds — no model
changes required.

**Extensibility:** to include a case from Group B-1 (exposure Oct 7–25,
onset day $O$), set `T_lower = O − 25 − 1`, `T_upper = O − 7 + 1`.

------------------------------------------------------------------------

### 3.2 Sampling

```{r stan-sample, cache=TRUE, results='hide'}
# Derive per-observation bounds from integer incubation periods.
# Group A: single calendar-day exposure and onset -> bounds = k +/- 1.
T_lower <- as.numeric(inc) - 1.0
T_upper <- as.numeric(inc) + 1.0

stan_data <- list(
  N       = length(inc),
  T_lower = T_lower,
  T_upper = T_upper
)

fit_stan <- stan(
  model_code = stan_code,
  data       = stan_data,
  chains     = 4,
  iter       = 10000,
  warmup     = 5000,
  seed       = 42,
  refresh    = 0
)
```

4 chains × 5,000 post-warmup iterations = 20,000 posterior samples.

------------------------------------------------------------------------

### 3.3 HMC Diagnostics

```{r stan-diagnostics, message=TRUE}
check_hmc_diagnostics(fit_stan)

stan_sum <- summary(
  fit_stan, pars = c("mean_ip", "sd_ip", "mu", "sigma")
)$summary
diag_df <- data.frame(
  Parameter = c(
    "mean_ip (days)", "sd_ip (days)",
    "mu (meanlog)", "sigma (sdlog)"
  ),
  Mean  = round(stan_sum[, "mean"], 4),
  R_hat = round(stan_sum[, "Rhat"], 4),
  ESS   = round(stan_sum[, "n_eff"])
)
kable(
  diag_df,
  col.names = c("Parameter", "Posterior Mean", "R-hat", "ESS"),
  row.names = FALSE
)
```

**Interpretation:**

- **R-hat < 1.01** for all parameters confirms convergence across chains.
- **ESS** — effective sample size; target > 400.
- **Divergent transitions** — zero is expected; any divergence requires
  investigation.
- **E-BFMI** — values near 1 indicate efficient posterior exploration.

------------------------------------------------------------------------

### 3.4 Trace Plots

```{r trace, fig.width=10, fig.height=6}
chains_arr <- extract(
  fit_stan, pars = c("mean_ip", "sd_ip"),
  permuted = FALSE, inc_warmup = FALSE
)

df_trace <- do.call(rbind, lapply(seq_len(dim(chains_arr)[2]), function(k) {
  data.frame(
    iter    = seq_len(dim(chains_arr)[1]),
    mean_ip = chains_arr[, k, "mean_ip"],
    sd_ip   = chains_arr[, k, "sd_ip"],
    chain   = factor(k)
  )
}))

p_tr_mean <- ggplot(
  df_trace, aes(x = iter, y = mean_ip, color = chain)
) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  labs(
    title = "Trace — mean_ip (days) [post-warmup]",
    x = "post-warmup iteration", y = "mean_ip"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p_tr_sd <- ggplot(
  df_trace, aes(x = iter, y = sd_ip, color = chain)
) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  labs(
    title = "Trace — sd_ip (days) [post-warmup]",
    x = "post-warmup iteration", y = "sd_ip"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p_tr_mean, p_tr_sd, ncol = 1)
```

------------------------------------------------------------------------

### 3.5 Posterior Distributions

```{r posterior, fig.width=10, fig.height=4}
post_stan <- as.data.frame(
  extract(fit_stan, pars = c("mean_ip", "sd_ip", "mu", "sigma"))
)

mle_mean <- exp(fit_lnorm$par[1] + fit_lnorm$par[2]^2 / 2)
mle_sd <- sqrt(
  (exp(fit_lnorm$par[2]^2) - 1) *
    exp(2 * fit_lnorm$par[1] + fit_lnorm$par[2]^2)
)

p_post_mean <- ggplot(post_stan, aes(x = mean_ip)) +
  geom_density(fill = "#009E73", alpha = 0.5) +
  geom_vline(
    xintercept = mle_mean, linetype = "dashed", color = "gray30"
  ) +
  labs(
    title    = "Posterior — mean_ip (days)",
    subtitle = "Dashed: MLE-equivalent mean (point likelihood)",
    x = "mean incubation period (days)", y = "density"
  ) +
  theme_minimal()

p_post_sd <- ggplot(post_stan, aes(x = sd_ip)) +
  geom_density(fill = "#009E73", alpha = 0.5) +
  geom_vline(
    xintercept = mle_sd, linetype = "dashed", color = "gray30"
  ) +
  labs(
    title    = "Posterior — sd_ip (days)",
    subtitle = "Dashed: MLE-equivalent SD (point likelihood)",
    x = "SD incubation period (days)", y = "density"
  ) +
  theme_minimal()

grid.arrange(p_post_mean, p_post_sd, ncol = 2)
```

**Interpretation:**

- **mean_ip** — posterior concentrated ~20–24 days; the MLE reference
  line confirms agreement between the prior and the interval-censored
  likelihood.

- **sd_ip** — right-skewed; mode ~3 days, reflecting individual-level
  variability in incubation length.

------------------------------------------------------------------------

### 3.6 Posterior Predictive Check

```{r posterior-predictive, fig.width=8, fig.height=5}
set.seed(1)
n_samp   <- 500
samp_idx <- sample(nrow(post_stan), n_samp)
pp_mat <- sapply(samp_idx, function(i) {
  dlnorm(freq, post_stan$mu[i], post_stan$sigma[i])
})

df_pp <- data.frame(
  x     = freq,
  mean  = rowMeans(pp_mat),
  lower = apply(pp_mat, 1, quantile, probs = 0.025),
  upper = apply(pp_mat, 1, quantile, probs = 0.975)
)

ggplot(df_pp, aes(x = x)) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper), fill = "#009E73", alpha = 0.25
  ) +
  geom_line(aes(y = mean), color = "#009E73", linewidth = 1) +
  geom_rug(
    data = data.frame(x = inc), aes(x = x),
    inherit.aes = FALSE, sides = "b", alpha = 0.8
  ) +
  labs(
    title    = "Posterior Predictive — Lognormal (Stan/HMC)",
    subtitle = paste0(
      "Mean density +/- 95% interval from ", n_samp,
      " posterior samples"
    ),
    x = "days", y = "density"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## 4. Final Estimates

```{r final-estimates}
est_df <- data.frame(
  Quantity = c(
    "Mean incubation period (days)",
    "SD incubation period (days)",
    "Median incubation period (days)",
    "mu (meanlog)",
    "sigma (sdlog)"
  ),
  Posterior_Median = round(c(
    median(post_stan$mean_ip),
    median(post_stan$sd_ip),
    median(exp(post_stan$mu)),
    median(post_stan$mu),
    median(post_stan$sigma)
  ), 2),
  CI_Lower = round(c(
    quantile(post_stan$mean_ip, 0.025),
    quantile(post_stan$sd_ip,   0.025),
    quantile(exp(post_stan$mu), 0.025),
    quantile(post_stan$mu,      0.025),
    quantile(post_stan$sigma,   0.025)
  ), 2),
  CI_Upper = round(c(
    quantile(post_stan$mean_ip, 0.975),
    quantile(post_stan$sd_ip,   0.975),
    quantile(exp(post_stan$mu), 0.975),
    quantile(post_stan$mu,      0.975),
    quantile(post_stan$sigma,   0.975)
  ), 2)
)
kable(
  est_df,
  col.names = c(
    "Quantity", "Posterior Median", "95% CrI Lower", "95% CrI Upper"
  ),
  row.names = FALSE
)
```

**Interpretation:**

The posterior median and 95% credible interval give point estimates
usable for outbreak planning. The interval-censored framework is
directly extensible to cases with wider exposure windows (Groups B–E)
should individual-level $[E_L, E_R]$ and $[O_L, O_R]$ data become
available — no changes to the Stan model would be required.

------------------------------------------------------------------------

## 5. Credits

### Data & Literature

- Kulagin, S.V., Fedorova, N.I., & Ketiladze, E.S. (1962). Laboratory
  outbreak of hemorrhagic fever with renal syndrome. *Journal of
  Microbiology, Epidemiology and Immunology*, 33(10), 121–126.

- Sartwell, P.E. (1950). The distribution of incubation periods of
  infectious disease. *American Journal of Hygiene*, 51(3), 310–318.

- Reich, N.G., et al. (2009). Estimating incubation period distributions
  with coarse data. *Statistics in Medicine*, 28(22), 2769–2784.
  *(Foundational reference for the doubly interval-censored framework.)*

### Software

- R Core Team. R: A Language and Environment for Statistical Computing.
  R Foundation for Statistical Computing, Vienna, Austria.
- Stan Development Team. RStan: the R interface to Stan.
- Wickham, H. ggplot2: Elegant Graphics for Data Analysis.
  Springer-Verlag New York, 2016.
