---
title: "Internship at TWCR"
author: "Andy Chu"
date: "2024-08-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TWCR
https://twcr.tw/ 


TWCR, short for Taiwan Cancer Registry Center, is a data center located in the Public Health building of NTU.
It has been requested by the government to collect and sort cancer data, to build a national cancer registry database.
Due to the high coverage of health insurance in Taiwan, TWCR is able to possess a large amount of high-quality data.
They publish annual reports on cancer in Taiwan, using the database.

The reports can be found in:

https://twcr.tw/?page_id=1354
https://www.hpa.gov.tw/Pages/List.aspx?nodeid=269

## Heatmap

However, these annual reports consist of countless descriptive charts and text.
I proposed a way to visualize the data, inspired by the GBD Compare website from IHME.
https://vizhub.healthdata.org/gbd-compare/

Using a heatmap, people will be able to browse through the whole dataset rapidly, while having a clear picture of how the incidence and mortality rank.

Note: the original analysis used TWCR's internal registry data, which cannot leave
the center. The table below contains pseudo-data reconstructed from the exported figure
(same rank values, same layout) purely to illustrate the visualization method.

```{r heatmap-setup, include=FALSE}
# Namespaced calls only (no library() attach) so this page's rendering
# can't be affected by functions masked by other pages' packages when
# the whole site is built in one shared R session.
```

## Pseudo Data (Hard-coded)
```{r heatmap-data}
# Pseudo rank data (22 counties x 11 cancer sites, 1 = highest incidence rank)
rank_wide <- tibble::tribble(
  ~county,   ~全癌症, ~女性乳房, ~肺, ~攝護腺, ~結直腸, ~肝臟, ~口腔, ~子宮, ~甲狀腺, ~卵巢, ~胃,
  "高雄市",  1,  10, 13,  7,  2,  3,  9,  6,  3, 12,  9,
  "台南市",  2,   8,  1,  5,  1,  5, 13,  7,  5,  7, 15,
  "台中市",  3,   4, 10,  2,  3,  7, 12, 10,  7,  2,  6,
  "新北市",  4,   1,  7,  3,  7, 17, 15,  2,  6, 10,  5,
  "嘉義市",  5,   5,  6,  4,  9,  9, 10, 16, 14,  1, 16,
  "屏東縣",  6,  16, 19, 13,  6,  3,  2,  3, 13,  6, 19,
  "宜蘭縣",  7,   9,  4, 14, 11,  8, 11, 11,  9, 20,  4,
  "新竹市",  8,   3,  9,  6, 18, 19, 16,  5,  1, 14, 12,
  "嘉義縣",  9,  14,  8, 17, 14,  2,  3, 12, 18, 15, 21,
  "苗栗縣", 10,  17, 16, 12,  5, 12,  8,  8,  8, 16, 13,
  "台北市", 11,   2,  3,  1, 12, 22, 20,  1,  4, 11, 10,
  "彰化縣", 12,  11, 12, 16,  4, 10,  7, 17, 11, 12, 11,
  "南投縣", 13,  13, 18, 11, 19, 14,  4, 15, 16,  3,  7,
  "雲林縣", 14,  19,  5,  8, 15,  1,  5, 14, 19, 19, 20,
  "桃園市", 15,   6, 14, 10, 10, 18, 14,  4, 12,  8, 14,
  "基隆市", 16,  12, 15,  9,  8, 15, 17, 13, 17,  9,  3,
  "花蓮縣", 17,  15, 20, 19, 17, 11,  6, 18, 21,  5,  1,
  "新竹縣", 18,   7, 17, 14, 13, 20, 18,  9,  2,  4, 17,
  "台東縣", 19,  21, 11, 18, 16,  6,  1, 19, 10, 18,  2,
  "澎湖縣", 20,  20,  2, 20, 20, 16, 19, 21, 15, 21, 22,
  "連江縣", 21,  18, 21, 22, 22, 13, 21, NA, 22, 17,  8,
  "金門縣", 22,  22, 22, 21, 21, 21, 22, 20, 20, 22, 18
)

cancer_levels <- c("全癌症", "女性乳房", "肺", "攝護腺", "結直腸",
                    "肝臟", "口腔", "子宮", "甲狀腺", "卵巢", "胃")

# Rank -> within-column percentile (rank 1 -> 100, largest rank in the
# column -> 0). 連江縣's 子宮 (uterine) rank is NA: the female population
# there is too small to compute a stable rank, shown as grey in the plot.
rank_long <- rank_wide |>
  tidyr::pivot_longer(cols = -county, names_to = "cancer", values_to = "rank") |>
  dplyr::mutate(cancer = factor(cancer, levels = cancer_levels)) |>
  dplyr::group_by(cancer) |>
  dplyr::mutate(
    n_valid = sum(!is.na(rank)),
    percentile = 100 * (n_valid - rank) / (n_valid - 1)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    county = factor(county, levels = rev(rank_wide$county)),
    label_color = dplyr::if_else(
      percentile >= 65 | percentile <= 20, "white", "black"
    )
  )
```


## Output

The ranking of incidence for the top 10 cancers in Taiwan.

The x-axis represents the cancer site (tumor location); the y-axis represents the city/county.
```{r heatmap-plot, fig.width=9, fig.height=8, warning=FALSE}
ggplot2::ggplot(
  rank_long, ggplot2::aes(x = cancer, y = county, fill = percentile)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.6) +
  ggplot2::geom_text(
    ggplot2::aes(label = rank, color = label_color),
    size = 4, show.legend = FALSE
  ) +
  ggplot2::scale_color_identity() +
  ggplot2::scale_fill_distiller(
    palette = "RdYlBu", direction = -1,
    limits = c(0, 100), na.value = "grey80",
    name = "百分位數", breaks = c(0, 50, 100)
  ) +
  ggplot2::scale_x_discrete(position = "bottom") +
  ggplot2::labs(
    title = "各縣市十大癌症發生率排名",
    x = "發生部位", y = NULL
  ) +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(angle = 60, hjust = 1),
    plot.title = ggplot2::element_text(hjust = 0.5, face = "bold")
  )
```
