---
title: "2011 Listeria Outbreak"
subtitle: "Early-Stage Outbreak Forecasting with GMITS"
author: "Andy Chu"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
for (pkg in c("tidyverse", "kableExtra")) {
  suppressMessages(
    library(pkg, character.only = TRUE, quietly = TRUE)
  )
}
```

## Functions {-}

### mae(), mape() {-}

```{r fn-errors}
mae <- function(actual, predicted) {
  mean(abs(actual - predicted), na.rm = TRUE)
}

mape <- function(actual, predicted) {
  valid <- actual != 0
  mean(
    abs((actual[valid] - predicted[valid]) / actual[valid]),
    na.rm = TRUE
  ) * 100
}
```

### gmits() {-}

```{r fn-gmits}
# n_ahead: out-of-sample steps using mean(delta_t) as step size.
# Returns $x0_future (and $t_future as numeric time offsets) when n_ahead > 0.
gmits <- function(count, date, n_ahead = 0) {
  valid <- !is.na(count) & !is.na(date)
  x0    <- count[valid]
  date  <- date[valid]
  if (length(x0) < 2) {
    warning("Not enough valid data points for modeling.")
    return(NULL)
  }
  if (x0[1] == 0) x0[1] <- 1e-3
  n <- length(x0)

  delta_t    <- as.numeric(
    difftime(date[2:n], date[1:(n - 1)], units = "days")
  )
  time_index <- cumsum(c(0, delta_t))

  x1 <- cumsum(c(x0[1], x0[2:n] * delta_t))
  z  <- 0.5 * (x1[2:n] + x1[1:(n - 1)])

  B <- cbind(-z * delta_t, delta_t)
  Y <- x0[2:n]
  W <- diag(delta_t)
  p <- solve(t(B) %*% W %*% B) %*% t(B) %*% W %*% Y
  a_hat <- p[1]
  b_hat <- p[2]

  x1_hat <- c(
    x1[1],
    (x1[1] - b_hat / a_hat) *
      exp(-a_hat * (time_index[-1] - time_index[1])) +
      b_hat / a_hat
  )
  x0_hat <- c(x0[1], diff(x1_hat) / delta_t)

  result <- list(
    a = a_hat, b = b_hat,
    x1_hat = x1_hat, x0_hat = x0_hat,
    date = date, delta_t = delta_t
  )

  if (n_ahead > 0) {
    dt_mean  <- mean(delta_t)
    t_future <- time_index[n] + cumsum(rep(dt_mean, n_ahead))
    x1_fut   <- (x1[1] - b_hat / a_hat) *
      exp(-a_hat * (t_future - time_index[1])) + b_hat / a_hat
    result$x0_future   <- c(
      (x1_fut[1] - x1_hat[n]) / dt_mean,
      diff(x1_fut) / dt_mean
    )
    result$date_future <- date[n] + round(dt_mean * seq_len(n_ahead))
  }
  result
}
```

### gmits_boot() {-}

```{r fn-bootstrap}
# n_ahead > 0: point/lower/upper each extend to n + n_ahead;
# last n_ahead entries are the out-of-sample prediction interval.
gmits_boot <- function(count, date, n_boot = 1000, ci = 0.95, n_ahead = 0) {
  base <- gmits(count, date, n_ahead = n_ahead)
  if (is.null(base)) return(NULL)

  valid      <- !is.na(count) & !is.na(date)
  x0         <- count[valid]
  dv         <- date[valid]
  n          <- length(x0)
  resid_pool <- (x0 - base$x0_hat)[2:n]
  n_out      <- n + n_ahead

  boot_mat <- replicate(n_boot, {
    e_boot  <- sample(resid_pool, n - 1, replace = TRUE)
    x0_boot <- c(x0[1], pmax(base$x0_hat[2:n] + e_boot, 0))
    res     <- tryCatch(
      gmits(x0_boot, dv, n_ahead = n_ahead),
      error = function(e) NULL
    )
    if (is.null(res)) return(rep(NA_real_, n_out))
    c(res$x0_hat, if (n_ahead > 0) res$x0_future else NULL)
  })

  alpha <- (1 - ci) / 2
  list(
    point = c(base$x0_hat,
              if (n_ahead > 0) base$x0_future else NULL),
    lower = apply(boot_mat, 1, quantile, probs = alpha,     na.rm = TRUE),
    upper = apply(boot_mat, 1, quantile, probs = 1 - alpha, na.rm = TRUE),
    a = base$a,
    b = base$b
  )
}
```

### exp_reg() {-}

```{r fn-expreg}
# Log-linear OLS baseline: log(y) ~ t.
# Only non-zero observations are used (consistent with gmits non-zero day approach).
# n_ahead: step forward using mean(delta_t) of the input dates.
exp_reg <- function(count, date, n_ahead = 0) {
  valid <- !is.na(count) & count > 0 & !is.na(date)
  x0 <- count[valid]
  dv <- date[valid]
  t_num <- as.numeric(dv - dv[1])
  fit <- lm(log(x0) ~ t_num)
  n <- length(x0)
  dt_mean <- if (n > 1) mean(diff(t_num)) else 1

  result <- list(
    alpha    = coef(fit)[["(Intercept)"]],
    beta     = coef(fit)[["t_num"]],
    x0_hat   = exp(fitted(fit)),
    date     = dv
  )
  if (n_ahead > 0) {
    t_future <- t_num[n] + cumsum(rep(dt_mean, n_ahead))
    result$x0_future   <- exp(
      predict(fit, newdata = data.frame(t_num = t_future))
    )
    result$date_future <- dv[n] + round(dt_mean * seq_len(n_ahead))
  }
  result
}
```

### truncation_adjust() {-}

```{r fn-truncation}
# Multiply the most recent n_recent observations by scale_factor
# before passing to gmits(), simulating upward correction of right-truncated data.
truncation_adjust <- function(count, n_recent = 1, scale_factor = 1.5) {
  adj <- count
  idx <- tail(seq_along(count), n_recent)
  adj[idx] <- count[idx] * scale_factor
  adj
}
```

---

# Synthetic Validation

Validates that GMITS recovers known parameters from irregular-interval data with
no noise. Two scenarios: growth phase (a < 0) and decay phase (a > 0).

```{r synth-validate}
# Generate x0 from ODE solution with known (a, b, x0[1]) and irregular spacing
synth_series <- function(a, b, x0_1, t_pts) {
  x1 <- (x0_1 - b / a) * exp(-a * t_pts) + b / a
  dt <- diff(t_pts)
  x0 <- c(x0_1, diff(x1) / dt)
  list(x0 = x0, dates = as.Date("2024-01-01") + t_pts)
}

t_irr    <- c(0, 2, 4, 7, 10)   # irregular gaps: 2, 2, 3, 3
n_ahead_s <- 3

# Growth scenario
s1 <- synth_series(a = -0.4, b = 3, x0_1 = 2, t_pts = t_irr)
r1 <- gmits(s1$x0, s1$dates, n_ahead = n_ahead_s)

# Decay scenario
s2 <- synth_series(a = 0.3, b = 5, x0_1 = 8, t_pts = t_irr)
r2 <- gmits(s2$x0, s2$dates, n_ahead = n_ahead_s)

# True ODE continuation for comparison in plot
true_oos <- function(a, b, x0_1, t_fit, n_ahead) {
  dt_mean  <- mean(diff(t_fit))
  t_fut    <- t_fit[length(t_fit)] + cumsum(rep(dt_mean, n_ahead))
  t_all    <- c(t_fit, t_fut)
  x1_all   <- (x0_1 - b / a) * exp(-a * t_all) + b / a
  dt_all   <- diff(t_all)
  c(x0_1, diff(x1_all) / dt_all)
}

synth_true1 <- true_oos(-0.4, 3, 2, t_irr, n_ahead_s)
synth_true2 <- true_oos( 0.3, 5, 8, t_irr, n_ahead_s)

data.frame(
  Scenario    = c("Growth (a = -0.4)", "Decay (a = 0.3)"),
  a_true      = c(-0.4,  0.3),
  b_true      = c( 3.0,  5.0),
  a_hat       = round(c(r1$a, r2$a), 6),
  b_hat       = round(c(r1$b, r2$b), 6),
  MAE_in_samp = c(
    mae(s1$x0, r1$x0_hat),
    mae(s2$x0, r2$x0_hat)
  )
) |>
  kable(
    digits  = 6,
    caption = "GMITS parameter recovery on noise-free synthetic data"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r plot-synth-validate}
n_fit_s <- length(t_irr)
dt_mean_s <- mean(diff(t_irr))
t_oos   <- t_irr[n_fit_s] + cumsum(rep(dt_mean_s, n_ahead_s))
t_all_s <- c(t_irr, t_oos)

make_synth_df <- function(scenario, true_vals, fitted_vals, future_vals,
                          true_oos_vals) {
  n_fit <- length(true_vals)
  data.frame(
    t        = t_all_s,
    True     = c(true_vals, true_oos_vals),
    Fitted   = c(fitted_vals, future_vals),
    Forecast = c(rep(FALSE, n_fit), rep(TRUE, n_ahead_s)),
    Scenario = scenario
  )
}

df_synth_plot <- bind_rows(
  make_synth_df(
    "Growth (a = -0.4)", s1$x0, r1$x0_hat,
    r1$x0_future, tail(synth_true1, n_ahead_s)
  ),
  make_synth_df(
    "Decay (a = 0.3)", s2$x0, r2$x0_hat,
    r2$x0_future, tail(synth_true2, n_ahead_s)
  )
) |>
  pivot_longer(c(True, Fitted), names_to = "type", values_to = "value")

ggplot(df_synth_plot, aes(t, value, color = type, linetype = Forecast)) +
  geom_vline(
    xintercept = t_irr[n_fit_s] + dt_mean_s / 2,
    linetype = "dotted", color = "grey60"
  ) +
  geom_line(linewidth = 0.9) +
  geom_point(
    data = df_synth_plot |> filter(!Forecast, type == "True"),
    size = 2.5
  ) +
  scale_color_manual(
    values = c(True = "black", Fitted = "steelblue"),
    name   = NULL
  ) +
  scale_linetype_manual(
    values = c("FALSE" = "solid", "TRUE" = "dashed"),
    labels = c("FALSE" = "in-sample", "TRUE" = "forecast"),
    guide  = guide_legend(title = NULL)
  ) +
  facet_wrap(~ Scenario, scales = "free_y", ncol = 2) +
  theme_minimal() +
  labs(
    title    = "GMITS Synthetic Validation — Parameter Recovery",
    subtitle = "Black = true ODE; Blue = GMITS fit + 3-step forecast; Dotted = forecast boundary",
    x = "Time (days from t₀)", y = "Period incidence"
  )
```

**Observed limitation — growth phase forecast bias.**
The growth scenario (a = −0.4) shows systematic underprediction in the
forecast steps even under noise-free conditions, while the decay scenario
(a = 0.3) tracks the true series closely. This is not an implementation
error; it is an inherent approximation bias in the GM(1,1) family.

The background value $z_{k} = \frac{1}{2}(x^{(1)}_{k} + x^{(1)}_{k-1})$
approximates the interval mean of $x^{(1)}$ by the arithmetic mean of its
endpoints. For an exponentially growing $x^{(1)}$ (convex curve), the
arithmetic mean of the endpoints *understates* the true integral mean,
causing the model to estimate a slower growth rate than the truth. The bias
scales with $|a| \times \Delta t$: faster growth or wider reporting
intervals amplify it. In the decay scenario $x^{(1)}$ is concave, the bias
acts in the opposite direction, and the absolute magnitude is smaller.

This limitation is well-known in the grey system literature (Liu, 2025;
Liu & Deng, 2000) and is the motivation for optimising the background
coefficient $\alpha$ (replacing the fixed 0.5). That extension is outside
the scope of this study; the bias is reported here as an explicit boundary
condition for GMITS applicability: **the method is most reliable when
$|a| \times \overline{\Delta t}$ is small**, i.e., moderate growth rates
and short reporting intervals.

---

# Data

## 2011 Listeria Outbreak

Sporadic foodborne transmission; illness-onset series (A files).
Snapshot `091211A` (Sep 12, 2011) is the **input**; `082712A` (Aug 27, 2012)
is the **ground truth** (final report).

```{r data-listeria}
wd <- "d:/Projects/RA/Listeria/data mdy"

read_snap <- function(fname) {
  read_csv(
    file.path(wd, fname),
    col_types = cols(Date = col_character(), Count = col_integer())
  ) |>
    mutate(
      Date = as.Date(Date),
      Snapshot = str_remove(fname, "\\.csv$")
    )
}

lst_input <- read_snap("091211A.csv") |> rename(Count_input = Count)
lst_gt    <- read_snap("082712A.csv") |> rename(Count_gt    = Count)

lst_combined <- full_join(
  lst_input |> select(Date, Count_input),
  lst_gt    |> select(Date, Count_gt),
  by = "Date"
) |>
  arrange(Date)

# Non-zero reporting days from input snapshot (091211A)
lst_nonzero <- lst_input |>
  filter(Count_input > 0) |>
  arrange(Date)

# First 6 non-zero days (n=6 primary)
lst_n6 <- lst_nonzero |> slice_head(n = 6)
# First 5 non-zero days
lst_n5 <- lst_nonzero |> slice_head(n = 5)
# First 4 non-zero days
lst_n4 <- lst_nonzero |> slice_head(n = 4)
```

```{r listeria-input-table}
lst_n6 |>
  mutate(delta_t = c(NA, diff(as.numeric(Date)))) |>
  kable(
    col.names = c("Date", "Count (input)", "Snapshot", "Δt (days)"),
    caption   = "Listeria input: first 6 non-zero reporting days (091211A)"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Early-Stage Forecasting

## Listeria (n = 4 / 5 / 6)

Three input sizes are compared. n = 6 is the primary scenario (maximum data
available from the Sep 12 snapshot before the forecast window). n = 4 and
n = 5 demonstrate sensitivity to input size.

```{r forecast-listeria}
n_ahead_l <- 4

run_listeria_fc <- function(df_n) {
  res_g  <- gmits_boot(df_n$Count_input, df_n$Date,
                       n_boot = 1000, n_ahead = n_ahead_l)
  res_er <- exp_reg(df_n$Count_input, df_n$Date, n_ahead = n_ahead_l)
  n_fit  <- nrow(df_n)
  dt_mean <- mean(diff(as.numeric(df_n$Date)))
  data.frame(
    Date     = c(df_n$Date,
                 df_n$Date[n_fit] + round(dt_mean * seq_len(n_ahead_l))),
    Point    = res_g$point,
    Lower    = res_g$lower,
    Upper    = res_g$upper,
    ExpReg   = c(res_er$x0_hat, res_er$x0_future),
    Forecast = c(rep(FALSE, n_fit), rep(TRUE, n_ahead_l)),
    n_label  = paste0("n = ", n_fit)
  )
}

df_lst_fc <- bind_rows(
  run_listeria_fc(lst_n4),
  run_listeria_fc(lst_n5),
  run_listeria_fc(lst_n6)
) |>
  mutate(n_label = factor(n_label, levels = c("n = 4", "n = 5", "n = 6")))
```

```{r plot-listeria-forecast}
gt_l <- lst_combined |>
  filter(!is.na(Count_gt)) |>
  select(Date, GT = Count_gt)

ggplot() +
  geom_col(
    data = gt_l |>
      filter(Date >= as.Date("2011-08-16"),
             Date <= as.Date("2011-09-10")),
    aes(Date, GT),
    fill = "grey80", alpha = 0.6
  ) +
  geom_ribbon(
    data = df_lst_fc,
    aes(Date, ymin = Lower, ymax = Upper, fill = n_label),
    alpha = 0.2
  ) +
  geom_line(
    data = df_lst_fc,
    aes(Date, Point, color = n_label, linetype = Forecast),
    linewidth = 0.9
  ) +
  geom_line(
    data = df_lst_fc,
    aes(Date, ExpReg, color = n_label, linetype = Forecast),
    linetype = "dotted", linewidth = 0.7
  ) +
  scale_linetype_manual(
    values = c("FALSE" = "solid", "TRUE" = "dashed"),
    labels = c("FALSE" = "in-sample", "TRUE" = "forecast"),
    guide  = guide_legend(title = NULL)
  ) +
  scale_color_brewer(palette = "Dark2", name = "Input size") +
  scale_fill_brewer(palette  = "Dark2", name = "Input size") +
  scale_x_date(date_labels = "%m/%d", date_breaks = "7 days") +
  facet_wrap(~ n_label, ncol = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title    = "Listeria — GMITS 4-Step Forecast by Input Size",
    subtitle = "Solid/dashed = GMITS; Dotted = exp. regression; Grey = final ground truth (082712A)",
    x = "Date", y = "Cases"
  )
```

---

# Right Truncation Sensitivity

The most recent observation in the n = 4 input (Aug 23, count = 1) may be
underreported due to reporting delay. We scale it by 1.5× and 2.0× and
compare the resulting 4-step forecasts.

```{r truncation-sensitivity}
adj_none <- lst_n4$Count_input
adj_1_5  <- truncation_adjust(lst_n4$Count_input, n_recent = 1, scale_factor = 1.5)
adj_2_0  <- truncation_adjust(lst_n4$Count_input, n_recent = 1, scale_factor = 2.0)

run_trunc <- function(counts, label) {
  res    <- gmits_boot(counts, lst_n4$Date, n_boot = 1000, n_ahead = n_ahead_l)
  n_fit  <- nrow(lst_n4)
  dt_mean <- mean(diff(as.numeric(lst_n4$Date)))
  data.frame(
    Date      = c(lst_n4$Date,
                  lst_n4$Date[n_fit] + round(dt_mean * seq_len(n_ahead_l))),
    Point     = res$point,
    Lower     = res$lower,
    Upper     = res$upper,
    Forecast  = c(rep(FALSE, n_fit), rep(TRUE, n_ahead_l)),
    Adjustment = label
  )
}

df_trunc <- bind_rows(
  run_trunc(adj_none, "No adjustment"),
  run_trunc(adj_1_5,  "Scale ×1.5"),
  run_trunc(adj_2_0,  "Scale ×2.0")
) |>
  mutate(Adjustment = factor(
    Adjustment,
    levels = c("No adjustment", "Scale ×1.5", "Scale ×2.0")
  ))
```

```{r plot-truncation}
ggplot() +
  geom_col(
    data = gt_l |>
      filter(Date >= as.Date("2011-08-16"),
             Date <= as.Date("2011-09-10")),
    aes(Date, GT),
    fill = "grey80", alpha = 0.6
  ) +
  geom_ribbon(
    data = df_trunc,
    aes(Date, ymin = Lower, ymax = Upper, fill = Adjustment),
    alpha = 0.2
  ) +
  geom_line(
    data = df_trunc,
    aes(Date, Point, color = Adjustment, linetype = Forecast),
    linewidth = 0.9
  ) +
  scale_linetype_manual(
    values = c("FALSE" = "solid", "TRUE" = "dashed"),
    guide  = guide_legend(title = NULL)
  ) +
  scale_color_manual(
    values = c("No adjustment" = "#1b9e77",
               "Scale ×1.5"   = "#d95f02",
               "Scale ×2.0"   = "#7570b3"),
    name = "Truncation adj."
  ) +
  scale_fill_manual(
    values = c("No adjustment" = "#1b9e77",
               "Scale ×1.5"   = "#d95f02",
               "Scale ×2.0"   = "#7570b3"),
    name = "Truncation adj."
  ) +
  scale_x_date(date_labels = "%m/%d", date_breaks = "7 days") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title    = "Listeria n = 4 — Right Truncation Sensitivity",
    subtitle = "Point + 95% PI under three adjustments; Grey = final ground truth",
    x = "Date", y = "Cases"
  )
```

---

# Out-of-Sample Error Summary

Ground truth: `082712A` at all 4 forecast steps (n = 6 input scenario).

```{r error-summary}
fc_n6 <- df_lst_fc |>
  filter(n_label == "n = 6", Forecast)

lst_gt_fc <- lst_combined |>
  filter(Date %in% fc_n6$Date) |>
  select(Date, Count_gt)

lst_err <- fc_n6 |>
  left_join(lst_gt_fc, by = "Date") |>
  filter(!is.na(Count_gt))

data.frame(
  Model    = c("GMITS", "Exp. Reg."),
  n_input  = c(6L, 6L),
  n_steps  = c(nrow(lst_err), nrow(lst_err)),
  MAE_oos  = c(
    mae(lst_err$Count_gt, lst_err$Point),
    mae(lst_err$Count_gt, lst_err$ExpReg)
  ),
  MAPE_oos = c(
    mape(lst_err$Count_gt, lst_err$Point),
    mape(lst_err$Count_gt, lst_err$ExpReg)
  )
) |>
  kable(
    digits    = 2,
    col.names = c("Model", "n input", "Steps verified",
                  "MAE (oos)", "MAPE % (oos)"),
    caption   = "Out-of-sample error: GMITS vs. exponential regression (Listeria, n = 6)"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```
